---
title: "Prediction Model for Progression-Free Survival in Breast Cancer Patients"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# Setup

```{r packages}

library(tidyverse)   
library(caret)       
library(glmnet)      
library(mgcv)        # GAM
library(earth)       # MARS
library(corrplot)    
library(knitr)       
library(kableExtra)  
library(vip)         # variable importance
library(car)         

```

```{r seed-and-cv}

set.seed(1)

ctrl <- trainControl(
  method    = "repeatedcv",
  number    = 10,       # 10-fold
  repeats   = 5,        # 5 times
  savePredictions = TRUE
)
```

---

# Part I: Prediction Model for datA (1990–1995)

## Data Loading & Preprocessing

```{r load-data}

load("datA.RData")   

str(datA)

```

```{r preprocessing}

datA <- datA |>
  mutate(
    menopause       = factor(menopause, levels = c(0, 1),
                             labels = c("Pre", "Post")),
    race            = factor(race, levels = c(1, 2, 3, 4),
                             labels = c("White", "Black", "Asian", "Hispanic")),
    tumor_grade     = factor(tumor_grade, levels = c(1, 2, 3),
                             labels = c("Well", "Moderate", "Poor")),
    stage           = factor(stage, levels = c(1, 2, 3),
                             labels = c("I", "II", "III")),
    ER              = factor(ER,  levels = c(0, 1), labels = c("Neg", "Pos")),
    PR              = factor(PR,  levels = c(0, 1), labels = c("Neg", "Pos")),
    HER2            = factor(HER2, levels = c(0, 1), labels = c("Neg", "Pos")),
    surgery         = factor(surgery, levels = c(0, 1),
                             labels = c("BCS", "Mastectomy")),
    chemo           = factor(chemo, levels = c(0, 1), labels = c("No", "Yes")),
    hormonal_therapy = factor(hormonal_therapy, levels = c(0, 1),
                              labels = c("No", "Yes"))
  ) |>
  select(-any_of("id"))   # remove patient ID

# Reference levels: White, Grade Well, Stage I, Neg, BCS, No chemo, No hormonal
# (these are already the first levels set above)

dim(datA)

```

---

## Exploratory Data Analysis

```{r eda-setup}
cont_vars <- c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67", "log_PFS")
cat_vars  <- c("menopause", "race", "tumor_grade", "stage",
               "ER", "PR", "HER2", "surgery", "chemo", "hormonal_therapy")
```

### Outcome Distribution

```{r outcome-dist, fig.cap="Figure A1. Distribution of log(PFS) and normal Q-Q plot (datA)."}

p1 <- ggplot(datA, aes(x = log_PFS)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "steelblue", color = "white", alpha = 0.8) +
  geom_density(color = "darkred", linewidth = 0.8) +
  labs(x = "log(PFS) (months)", y = "Density") +
  theme_bw()

p2 <- ggplot(datA, aes(sample = log_PFS)) +
  stat_qq() + stat_qq_line(color = "darkred") +
  
  labs(x = "Theoretical", y = "Sample") +
  theme_bw()

gridExtra::grid.arrange(p1, p2, ncol = 2)

```

### Summary Statistics

```{r summary-continuous}
datA |>
  select(all_of(cont_vars)) |>
  summarise(across(everything(), list(
    Mean   = \(x) round(mean(x), 2),
    SD     = \(x) round(sd(x), 2),
    Median = \(x) round(median(x), 2),
    Min    = \(x) round(min(x), 2),
    Max    = \(x) round(max(x), 2)
  ))) |>
  pivot_longer(everything(), names_to = c("Variable", "Stat"), names_sep = "_(?=[^_]+$)") |>
  pivot_wider(names_from = Stat, values_from = value) |>
  kable(caption = "Table 1. Summary statistics of continuous variables (datA).") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 10)

```

```{r summary-categorical}

datA |>
  select(all_of(cat_vars)) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Level") |>
  count(Variable, Level) |>
  group_by(Variable) |>
  mutate(`n (%)` = paste0(n, " (", round(100 * n / sum(n), 1), "%)")) |>
  select(-n) |>
  kable(caption = "Table 2. Frequency distribution of categorical variables (datA).") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 10)

```

### Correlation Among Continuous Predictors

```{r corr-plot, fig.cap="Figure A2. Correlation matrix of continuous predictors (datA)."}

datA |>
  select(all_of(cont_vars)) |>
  cor() |>
  corrplot(method = "color", type = "upper",
           addCoef.col = "black", number.cex = 0.7, tl.cex = 0.8,
           mar = c(0, 0, 1, 0))

```

### Outcome by Predictors

```{r outcome-by-clinical}

p_stage <- ggplot(datA, aes(x = stage, y = log_PFS, fill = stage)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = "Stage", y = "log(PFS)", title = "By Stage") + theme_bw()

p_grade <- ggplot(datA, aes(x = tumor_grade, y = log_PFS, fill = tumor_grade)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = "Tumor Grade", y = "log(PFS)", title = "By Tumor Grade") + theme_bw()

p_er <- ggplot(datA, aes(x = ER, y = log_PFS, fill = ER)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = "ER Status", y = "log(PFS)", title = "By ER Status") + theme_bw()

p_her2 <- ggplot(datA, aes(x = HER2, y = log_PFS, fill = HER2)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = "HER2 Status", y = "log(PFS)", title = "By HER2 Status") + theme_bw()

gridExtra::grid.arrange(p_stage, p_grade, p_er, p_her2, ncol = 2)

```

```{r scatter-continuous}

datA |>
  select(log_PFS, age, bmi, tumor_size, lymph_nodes, Ki67) |>
  pivot_longer(-log_PFS, names_to = "Predictor", values_to = "Value") |>
  ggplot(aes(x = Value, y = log_PFS)) +
  geom_point(alpha = 0.15, size = 0.8, color = "steelblue") +
  geom_smooth(method = "loess", color = "darkred", se = TRUE, linewidth = 0.8) +
  facet_wrap(~Predictor, scales = "free_x", ncol = 3) +
  labs(x = NULL, y = "log(PFS)") +
  theme_bw()

```

### Summary

The outcome `log_PFS` is approximately normally distributed,
supporting the use of linear regression-based models.

Among continuous predictors, `lymph_nodes` (r = −0.32), `tumor_size` (r = −0.29), and `Ki67` (r = −0.25) show the strongest marginal associations with `log_PFS`, while `age`
and `bmi` show near-zero linear correlation. Critically, `tumor_size` and `lymph_nodes`are strongly collinear (r = 0.71), which inflates variance in OLS and motivates the use of **Elastic Net** regularization to stabilize coefficient estimates.

The LOESS plots reveal that `lymph_nodes` and `Ki67` exhibit non-linear relationships with
`log_PFS` — the effect of `lymph_nodes` appears to level off at higher values, and `Ki67`
shows a curved trend. This motivates **GAM** and **MARS** as non-linear alternatives.
`age` and `bmi` show relatively flat LOESS curves, suggesting their effects may be
conditional on other factors (e.g., menopausal status) rather than independently non-linear.

Together, these findings support training four candidate models: OLS (baseline), Elastic Net
(collinearity), GAM (non-linearity), and MARS (non-linearity + interactions).

---

## Model Training

### Model 1: OLS Linear Regression

```{r ols}

set.seed(1)
fit_ols <- train(
  log_PFS ~ .,
  data      = datA,
  method    = "lm",
  trControl = ctrl,
  preProcess = c("center", "scale")
)

summary(fit_ols$finalModel)

```

### Model 2: Elastic Net

```{r elastic-net}
set.seed(1)
enet_grid <- expand.grid(
  alpha  = seq(0, 1, by = 0.1),
  lambda = exp(seq(log(0.001), log(1), length.out = 50))
)

fit_enet <- train(
  log_PFS ~ .,
  data       = datA,
  method     = "glmnet",
  trControl  = ctrl,
  tuneGrid   = enet_grid,
  preProcess = c("center", "scale")
)

# Best tuning parameters
fit_enet$bestTune

# Plot CV-RMSE vs log(lambda), one line per alpha
fit_enet$results |>
  filter(!is.na(RMSE)) |>
  ggplot(aes(x = log(lambda), y = RMSE, color = factor(alpha), group = factor(alpha))) +
  geom_line() +
  labs(x = "log(lambda)", y = "CV-RMSE", color = "alpha",
       title = "Elastic Net: CV-RMSE by alpha and lambda") +
  theme_bw()

```

### Model 3: GAM

```{r gam}

set.seed(1)

# Final GAM on full datA
fit_gam <- mgcv::gam(
  log_PFS ~ s(age) + s(bmi) + s(tumor_size) + s(lymph_nodes) + s(Ki67) +
    menopause + race + tumor_grade + stage +
    ER + PR + HER2 + surgery + chemo + hormonal_therapy,
  data   = datA,
  method = "REML"
)

summary(fit_gam)

folds <- createFolds(datA$log_PFS, k = 10, list = TRUE, returnTrain = FALSE)

gam_preds <- numeric(nrow(datA))
for (i in seq_along(folds)) {
  train_i <- datA[-folds[[i]], ]
  test_i  <- datA[ folds[[i]], ]
  gam_i   <- mgcv::gam(
    log_PFS ~ s(age) + s(bmi) + s(tumor_size) + s(lymph_nodes) + s(Ki67) +
      menopause + race + tumor_grade + stage +
      ER + PR + HER2 + surgery + chemo + hormonal_therapy,
    data = train_i, method = "REML"
  )
  gam_preds[folds[[i]]] <- predict(gam_i, newdata = test_i)
}

gam_cv_rmse <- sqrt(mean((datA$log_PFS - gam_preds)^2))
gam_cv_r2   <- cor(datA$log_PFS, gam_preds)^2
cat("GAM 10-fold CV RMSE:", round(gam_cv_rmse, 4),
    " R2:", round(gam_cv_r2, 4), "\n")

```

### Model 4: MARS

```{r mars}

set.seed(1)
mars_grid <- expand.grid(
  degree = 1:2,
  nprune = 2:20
)

fit_mars <- train(
  log_PFS ~ .,
  data      = datA,
  method    = "earth",
  trControl = ctrl,
  tuneGrid  = mars_grid,
  preProcess = c("center", "scale")
)

fit_mars$bestTune
plot(fit_mars)

# Variable importance
vip::vip(fit_mars, num_features = 15)

```

---

## Model Comparison & Selection

```{r model-comparison}

# caret models compared via resamples()
resamples_list <- resamples(list(
  OLS        = fit_ols,
  ElasticNet = fit_enet,
  MARS       = fit_mars
))

cv_stats <- summary(resamples_list)$statistics

caret_rmse <- cv_stats$RMSE[, "Mean"]
caret_r2   <- cv_stats$Rsquared[, "Mean"]

# Add GAM 
all_cv <- data.frame(
  Model   = c(names(caret_rmse), "GAM"),
  CV_RMSE = round(c(caret_rmse, gam_cv_rmse), 4),
  CV_R2   = round(c(caret_r2,   gam_cv_r2),   4)
) |> arrange(CV_RMSE)

all_cv |>
  kable(caption = "Table 3. Cross-validated RMSE and R-squared for four candidate models (datA).",
        col.names = c("Model", "CV-RMSE", "CV-R2")) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 10)

```


### Final Model Selection

**Selected model: GAM**

```{r final-model-selection}

# GAM as final model
cat("Final model: GAM\n")
cat("CV-RMSE: ", round(gam_cv_rmse, 4), "\n")
cat("CV-R2:   ", round(gam_cv_r2,   4), "\n")

```

**Rationale.** Four candidate models were trained and compared using 10-fold
cross-validated RMSE (repeated 5 times): OLS (baseline), Elastic Net, MARS, and GAM.
All four models performed similarly (CV-RMSE range: 0.6975–0.7035), reflecting
the inherent predictive ceiling of 16 baseline clinical variables for individual-level
PFS prediction.

GAM was selected as the final model for the following reasons:

- **Best predictive performance**: GAM achieved the lowest CV-RMSE
- **Empirically justified non-linearity**: EDA identified non-linear LOESS patterns for
`bmi`, `tumor_size`, and `lymph_nodes`, validating the use of smooth terms over a purely linear specification.
- **Interactions not supported by data**: MARS with degree = 2 searched the full predictor
space for useful two-way interactions and selected degree = 1, indicating no interaction term
improved out-of-sample performance, supports an additive model structure.

The modest overall R-square (0.21) is consistent with clinical-only prediction models for
PFS in the literature and reflects unmeasured biological factors (e.g., genomic profiles)
rather than model misspecification.

---

## Final Model: Results & Interpretation

### Model Coefficients (Parametric Terms)

```{r gam-coef-table}
# Extract parametric coefficients from GAM
gam_coef <- summary(fit_gam)$p.table |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  rename(SE = `Std. Error`, `t-value` = `t value`, `p-value` = `Pr(>|t|)`)

gam_coef |>
  kable(caption = "Table 3a. GAM parametric coefficient estimates (datA).") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 10)
```

```{r gam-smooth-table}
# Smooth terms significance
gam_smooth <- summary(fit_gam)$s.table |>
  as.data.frame() |>
  rownames_to_column("Smooth Term") |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

gam_smooth |>
  kable(caption = "Table 3b. GAM smooth term significance (datA). edf > 1 indicates non-linearity.") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 10)
```

The following interpretation uses reference levels: White, pre-menopausal, Grade Well,
Stage I, ER/PR/HER2-negative, BCS, no chemotherapy, no hormonal therapy.

**Demographic characteristics:** Age showed a small but significant negative smooth effect
on log-PFS (p = 0.03), with older patients having slightly shorter PFS. BMI exhibited the
strongest non-linearity, suggesting a non-monotonic relationship with
PFS. Menopausal status and race were not significant after adjusting for other factors.

**Tumor characteristics:** Higher lymph node count was the strongest predictor of shorter
PFS (p < 0.001). Larger tumor size was associated with worse PFS (p = 0.006).
HER2-positive status was associated with 0.174 lower log-PFS compared to HER2-negative
(p < 0.001), reflecting its aggressive phenotype in this pre-trastuzumab era cohort.
ER-positive status was associated with 0.234 higher log-PFS (p < 0.001), consistent with
the favorable prognosis of hormone receptor-positive tumors. Poorly differentiated tumors
(Grade Poor) had significantly shorter PFS than well-differentiated tumors (est. = −0.162,
p = 0.006). PR, Ki67, and stage were not significant after adjustment.

**Treatment characteristics:** Hormonal therapy was the only treatment variable
significantly associated with PFS (est. = +0.133, p < 0.001), consistent with its
established efficacy for ER-positive disease. Surgery type and chemotherapy receipt were
not significant after adjusting for tumor and receptor characteristics.

### Partial Effect Plots (Smooth Terms)

```{r gam-partial-effects, fig.cap="Figure A5. Partial effects of continuous predictors in the final GAM (datA). Shaded bands are 95% credible intervals; rug marks show data density."}
par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
plot(fit_gam, shade = TRUE, shade.col = "lightblue",
     seWithMean = TRUE, scale = 0)
par(mfrow = c(1, 1))
```

### Performance Assessment

```{r gam-performance}
# Training RMSE and R2
train_pred  <- predict(fit_gam)
train_rmse  <- sqrt(mean((datA$log_PFS - train_pred)^2))
train_r2    <- summary(fit_gam)$r.sq

data.frame(
  Metric  = c("Training RMSE", "CV-RMSE (10-fold)", "Training R2 (adj)", "CV-R2"),
  Value   = round(c(train_rmse, gam_cv_rmse, train_r2, gam_cv_r2), 4)
) |>
  kable(caption = "Table 3c. GAM performance metrics (datA).") |>
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = FALSE, font_size = 10)
```



---

# Part II: Generalizability to datB (2000–2005)

## Cohort Comparison: datA vs datB

```{r load-datB}
load("datB.RData")

datB <- datB |>
  mutate(
    menopause        = factor(menopause, levels = c(0, 1), labels = c("Pre", "Post")),
    race             = factor(race, levels = c(1, 2, 3, 4),
                              labels = c("White", "Black", "Asian", "Hispanic")),
    tumor_grade      = factor(tumor_grade, levels = c(1, 2, 3),
                              labels = c("Well", "Moderate", "Poor")),
    stage            = factor(stage, levels = c(1, 2, 3), labels = c("I", "II", "III")),
    ER               = factor(ER,  levels = c(0, 1), labels = c("Neg", "Pos")),
    PR               = factor(PR,  levels = c(0, 1), labels = c("Neg", "Pos")),
    HER2             = factor(HER2, levels = c(0, 1), labels = c("Neg", "Pos")),
    surgery          = factor(surgery, levels = c(0, 1), labels = c("BCS", "Mastectomy")),
    chemo            = factor(chemo, levels = c(0, 1), labels = c("No", "Yes")),
    hormonal_therapy = factor(hormonal_therapy, levels = c(0, 1), labels = c("No", "Yes"))
  ) |>
  select(-any_of("id"))
```

```{r cohort-comparison}
# Outcome comparison
rbind(
  datA |> summarise(Cohort="datA (1990-95)", Mean=round(mean(log_PFS),3),
                    SD=round(sd(log_PFS),3), Median=round(median(log_PFS),3)),
  datB |> summarise(Cohort="datB (2000-05)", Mean=round(mean(log_PFS),3),
                    SD=round(sd(log_PFS),3), Median=round(median(log_PFS),3))
) |>
  kable(caption = "Table 4. Comparison of log(PFS) between cohorts.") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 10)

# t-test on outcome
t.test(datA$log_PFS, datB$log_PFS)
```

```{r continuous-comparison}
# Continuous predictor comparison
cont_pred <- c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")

map_dfr(cont_pred, function(v) {
  tibble(
    Variable = v,
    `datA mean (SD)` = paste0(round(mean(datA[[v]]), 2),
                               " (", round(sd(datA[[v]]), 2), ")"),
    `datB mean (SD)` = paste0(round(mean(datB[[v]]), 2),
                               " (", round(sd(datB[[v]]), 2), ")"),
    `Diff (B-A)`     = round(mean(datB[[v]]) - mean(datA[[v]]), 2),
    `p-value`        = round(t.test(datA[[v]], datB[[v]])$p.value, 4)
  )
}) |>
  kable(caption = "Table 5. Continuous predictor comparison between cohorts.") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 10)
```

```{r treatment-comparison}
# Treatment and key categorical comparison
treat_vars <- c("stage", "chemo", "hormonal_therapy", "surgery")

map_dfr(treat_vars, function(v) {
  pA <- round(100 * prop.table(table(datA[[v]])), 1)
  pB <- round(100 * prop.table(table(datB[[v]])), 1)
  tibble(
    Variable = v,
    `datA (%)` = paste(names(pA), pA, sep = "=", collapse = ", "),
    `datB (%)` = paste(names(pB), pB, sep = "=", collapse = ", ")
  )
}) |>
  kable(caption = "Table 6. Treatment and stage distribution between cohorts.") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 10)
```

```{r outcome-dist-comparison, fig.cap="Figure A6. Distribution of log(PFS) in datA vs datB."}
bind_rows(
  datA |> select(log_PFS) |> mutate(Cohort = "datA (1990-95)"),
  datB |> select(log_PFS) |> mutate(Cohort = "datB (2000-05)")
) |>
  ggplot(aes(x = log_PFS, fill = Cohort)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = c(mean(datA$log_PFS), mean(datB$log_PFS)),
             linetype = "dashed", color = c("steelblue", "coral")) +
  labs(x = "log(PFS) (months)", y = "Density") +
  theme_bw()
```


